---
title: "DataProcessing/Cleaning"
output: pdf_document
date: "2025-10-24"
---


Import required Libraries through the requirements.R file and import uncleaned data.
```{r setup, include=FALSE}
source("00_requirements.R")
uncleanData <- read_tsv("participants.tsv")
```

Changes column names to a more uniform syntaxing for easier reference and removes unused varaiables from the dataset.

```{r}

cleanData <- uncleanData[,!names(uncleanData) %in% c("HbA1c", "sex", "bodyweight")]
names(cleanData)[names(cleanData) == "fasting insulin"] <- "Fasting Insulin"
names(cleanData)[names(cleanData) == "age"] <- "Age"
names(cleanData)[names(cleanData) == "participant_id"] <- "Participant_ID"

print(cleanData)
```

HOMA is a score combining glucose + insulin. when HOMA is high that means bad glucose fuel(insulin resistance) which can relate to brain aging.

Hypothesis1: "if age is older then HOMA will be increase"

Create scatter plot to visualize the age relate to HOMA.

```{r}
plot(x = cleanData$Age , y = cleanData$HOMA , xlab = "age", ylab = "HOMA")

abline(h = 2.5, col = "red")

```
According to scatter plot of age and HOMA, we can observe HOMA is not really affected by age.
Red line indicates high HOMA; HOMA > 2.5 which means faster aging, yet data shows HOMA is not really related to age.


Hypothesis2: if BMI is  higher, does HOMA increases as well.

Create the scatter plot displays BMI with HOMA.

```{r}
plot(x = cleanData$BMI , y = cleanData$HOMA, xlab = "BMI", ylab = "HOMA")

abline(h = 2.5, col = "red")

```
Shows higher the BMI, HOMA tends to increase as well.
Red line indicates high HOMA. According to scatter plot, as BMI increases, HOMA increases as well.

high HOMA which can be high insulin resistance = glucose fuel less efficiently
- brain's activity is depends on glucose, its main fuel source.
- less efficient it is, there is high possibility that brain can aging faster.


